\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-05-24 doc/paper.Rmd
abstract

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- 
- 
bibliography: bibliography
abstract: Same-source analyses are a major part of a Forensic Toolmark Examiner's job. Technological advances, such as profilometers and confocal microscopy allow to measure 3D surfaces in a high-resolution digitized form. Methods for matching marks for a variety of tools have been studied in the literature. Analysis of these digitized markings require the use of statistical methods which can quantify the scientific mechanism of comparing markings and serve as basis for an error rate calculation. In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler} to assess striation marks on bullet lands for same-source identification. The Chumbley score  provides us with another approach in the same-source assessment of bullet striation marks. 
---

@@

<!--Bullet matching methods are usually based on these associated signatures.-->
In same-source comparisons this curvature is usually removed using some form of Gaussian filter \citep{ma2004} or non-parametric smoothing \citep{aoas}. 
\citet{chu2013} use an automatic method for counting consecutive matching striae (CMS). The authors report an error rate of 52% for known same-source land comparisons to be (incorrectly) identified as different-source (false negative) and zero false positives for known different-source lands. \citet{ma2004} and \citet{vorburger2011} discuss CCF (cross-correlation function) and its discriminating power and applicability for same-source analyses of bullets, but do not provide any error rates in their discussion. \citet{aoas} use multiple features, such as CCF, CMS, D (distance measure), etc. in a random forest based method and compare every land against every other land of  digitized versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST Ballistics Database \citep{nist}. The authors report an out-of-bag overall error rate of 0.46%, comprised of an error rate of 30.05% of  same-source pairs that were not identified and an error rate of 0.026% of different-source pairs that were incorrectly identified as same-source.

@@

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Method & Data Source & False Positives & False Negatives\\
  \hline
Consecutive Matching Bullets Striae (CMS) & & & \\ \textbf{\citet{aoas}} &  & 6.25\% &  33.851\% \\ \hline
Consecutive Non-matching Striae (CNMS) & Bullets & & \\  \textbf{\citet{aoas}} & & 6.26\% & 35.4166\%  \\ \hline
Average Distance (D) & Bullets  &  & \\ \textbf{\citet{aoas}} & & 6.25\% & 45.833\% \\ \hline 
Cross-correlation Function (CCF) & Bullets & & \\ \textbf{\citet{aoas}}  & & 6.25\% & 17.7083\%  \\ \hline
Sum of Peaks (S) & Bullets  & & \\  \textbf{\citet{aoas}} & & 6.25\% &  18.22915\% \\
   \hline
\end{tabular}}
\caption{\label{tab:bullets-ER} Error Rates in same-source bullet analysis reported in the literature.}
\end{table}

@@

\hh{What are we investigating in this paper, and what is the structure for the paper?}
\gk{In this paper we evaluate the adaptability of the chumbley score method to bullets, and in the process identify parameters that minimize the error rates. We then identify persistent scenarios in which the method proposed by \citet{hadler} fails to conduct a test. We go on to provide a solution to the failed tests problem, consequently increasing the power of the test and reducing error rates in the process. We also investigate the performance of the two algorithms using higher as well as lower resolution scans to understand if better scans and better equipment have a positive impact on the results.}

>>>>

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- land engraved areas (LEAs)
- algorithm
bibliography: bibliography
abstract: One of the major problems in forensic toolmark examination is the same-source problem, i.e. the identification of whether the marks on two different objects were made by the same tool. Profilometers and confocal microscopy and other technological advances allow us to measure 3D surfaces in previously unforeseen resolution and also enable digitized images. The literature provides various methods to measure similarity of toolmarks. In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler}, which was introduced for the assessment of marks made by screwdrivers, to assess striation marks on bullet lands for same-source identification. We provide methods to identify parameters that minimize the error rates for matching of LEAs at the example of the Hamby data, sets 44 and 252 measured by NIST and CSAFE. We provide a remedial algorithm to alleviate the problem of failed tests in the method proposed by \citet{hadler}. This increases the power of the test and reduces error rates in the process. Type II error rates for the Hamby data using the adjusted method showed an improvement of more than 20\% points over the performance of the algorithm proposed by \citet{hadler} (for a Type I error of 0.05). Overall, type II error rates of the adjusted method are on average at 0.22, putting the method on similar footing as other single feature matching approaches in the literature.
---

@@

<!--Bullet matching methods are usually based on these associated signatures.-->
In same-source comparisons this curvature is usually removed using some form of Gaussian filter \citep{ma2004} or non-parametric smoothing \citep{aoas}. An overview of some of the error rates reported in the literature on bullet matching is given in \autoref{tab:bullets-ER}.
\citet{chu2013} use an automatic method for counting consecutive matching striae (CMS). The authors report an error rate of 52% for known same-source land comparisons to be (incorrectly) identified as different-source (false negative) and zero false positives for known different-source lands. \citet{ma2004} and \citet{vorburger2011} discuss CCF (cross-correlation function) and its discriminating power and applicability for same-source analyses of bullets, but do not provide any error rates in their discussion. \citet{aoas} use multiple features, such as CCF, CMS, D (distance measure), etc. in a random forest based method and compare every land against every other land of  digitized versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST Ballistics Database \citep{nist}. The authors report an out-of-bag overall error rate of 0.46%, comprised of an error rate of 30.05% of  same-source pairs that were not identified and an error rate of 0.026% of different-source pairs that were incorrectly identified as same-source.

@@

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Method & Data Source & False Positives & False Negatives\\
  \hline
  \textbf{\citet{aoas}} & LEAs \\
Consecutive Matching Bullets Striae (CMS) &  &  6.25\% &  33.85\% \\ 
Consecutive Non-matching Striae (CNMS)  & &  6.26\% & 35.42\%  \\ 
Average Distance (D) &   &   6.25\% & 45.83\% \\ 
Cross-correlation Function (CCF) & &  6.25\% & 17.71\%  \\ 
Sum of Peaks (S) &  &  6.25\% &  18.23\% \\
   \hline
   \textbf{\citet{chu2013}} & LEAs \\
Consecutive Matching Bullets Striae (CMS) &  &  0\% &  52\% \\ 
\end{tabular}}
\caption{\label{tab:bullets-ER} Error Rates in same-source land-to-land analysis reported in the literature.}
\end{table}

@@

In this paper we evaluate the adaptability of the Chumbley score as a measure to quantify similarity in land engraved areas (LEAs) on bullets. For that we briefly introduce the deterministic method suggested by \citet{hadler} in section 2. In the process we provide methods to identify parameters that minimize the error rates. We then investigate persistent scenarios in which the method proposed by \citet{hadler} fails to come to a result. We go on to provide a solution to the failed tests problem, consequently increasing the power of the test and reducing error rates in the process. We set up a testing framework to compare the performance of the two algorithms in section 3 and finally discuss results in section 4.

### 2018-05-24 doc/paper.tex
abstract

\bigskip
\begin{abstract}
Same-source analyses are a major part of a Forensic Toolmark Examiner's
job. Technological advances, such as profilometers and confocal
microscopy allow to measure 3D surfaces in a high-resolution digitized
form. Methods for matching marks for a variety of tools have been
studied in the literature. Analysis of these digitized markings require
the use of statistical methods which can quantify the scientific
mechanism of comparing markings and serve as basis for an error rate
calculation. In this paper, we are investigating the applicability of
the Chumbley scoring method by \citet{hadler} to assess striation marks
on bullet lands for same-source identification. The Chumbley score
provides us with another approach in the same-source assessment of
bullet striation marks.
\end{abstract}

\noindent%
{\it Keywords:} forensic science, toolmark, cross-correlation, Mann-Whitney U statistic, , 
\vfill

@@

In same-source comparisons this curvature is usually removed using some
form of Gaussian filter \citep{ma2004} or non-parametric smoothing
\citep{aoas}. \citet{chu2013} use an automatic method for counting
consecutive matching striae (CMS). The authors report an error rate of
52\% for known same-source land comparisons to be (incorrectly)
identified as different-source (false negative) and zero false positives
for known different-source lands. \citet{ma2004} and
\citet{vorburger2011} discuss CCF (cross-correlation function) and its
discriminating power and applicability for same-source analyses of
bullets, but do not provide any error rates in their discussion.
\citet{aoas} use multiple features, such as CCF, CMS, D (distance
measure), etc. in a random forest based method and compare every land
against every other land of digitized versions of Hamby 252 and Hamby 44
\citep{hamby} published on the NIST Ballistics Database \citep{nist}.
The authors report an out-of-bag overall error rate of 0.46\%, comprised
of an error rate of 30.05\% of same-source pairs that were not
identified and an error rate of 0.026\% of different-source pairs that
were incorrectly identified as same-source.

@@

{\textcolor{orange}{What are we investigating in this paper, and what is the structure for the paper?}}

>>>>

\bigskip
\begin{abstract}
One of the major problems in forensic toolmark examination is the
same-source problem, i.e.~the identification of whether the marks on two
different objects were made by the same tool. Profilometers and confocal
microscopy and other technological advances allow us to measure 3D
surfaces in previously unforeseen resolution and also enable digitized
images. The literature provides various methods to measure similarity of
toolmarks. In this paper, we are investigating the applicability of the
Chumbley scoring method by \citet{hadler}, which was introduced for the
assessment of marks made by screwdrivers, to assess striation marks on
bullet lands for same-source identification. We provide methods to
identify parameters that minimize the error rates for matching of LEAs
at the example of the Hamby data, sets 44 and 252 measured by NIST and
CSAFE. We provide a remedial algorithm to alleviate the problem of
failed tests in the method proposed by \citet{hadler}. This increases
the power of the test and reduces error rates in the process. Type II
error rates for the Hamby data using the adjusted method showed an
improvement of more than 20\% points over the performance of the
algorithm proposed by \citet{hadler} (for a Type I error of 0.05).
Overall, type II error rates of the adjusted method are on average at
0.22, putting the method on similar footing as other single feature
matching approaches in the literature.
\end{abstract}

\noindent%
{\it Keywords:} forensic science, toolmark, cross-correlation, Mann-Whitney U statistic, land engraved areas (LEAs), algorithm
\vfill

@@

In same-source comparisons this curvature is usually removed using some
form of Gaussian filter \citep{ma2004} or non-parametric smoothing
\citep{aoas}. An overview of some of the error rates reported in the
literature on bullet matching is given in \autoref{tab:bullets-ER}.
\citet{chu2013} use an automatic method for counting consecutive
matching striae (CMS). The authors report an error rate of 52\% for
known same-source land comparisons to be (incorrectly) identified as
different-source (false negative) and zero false positives for known
different-source lands. \citet{ma2004} and \citet{vorburger2011} discuss
CCF (cross-correlation function) and its discriminating power and
applicability for same-source analyses of bullets, but do not provide
any error rates in their discussion. \citet{aoas} use multiple features,
such as CCF, CMS, D (distance measure), etc. in a random forest based
method and compare every land against every other land of digitized
versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST
Ballistics Database \citep{nist}. The authors report an out-of-bag
overall error rate of 0.46\%, comprised of an error rate of 30.05\% of
same-source pairs that were not identified and an error rate of 0.026\%
of different-source pairs that were incorrectly identified as
same-source.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Method & Data Source & False Positives & False Negatives\\
  \hline
  \textbf{\citet{aoas}} & LEAs \\
Consecutive Matching Bullets Striae (CMS) &  &  6.25\% &  33.85\% \\ 
Consecutive Non-matching Striae (CNMS)  & &  6.26\% & 35.42\%  \\ 
Average Distance (D) &   &   6.25\% & 45.83\% \\ 
Cross-correlation Function (CCF) & &  6.25\% & 17.71\%  \\ 
Sum of Peaks (S) &  &  6.25\% &  18.23\% \\
   \hline
   \textbf{\citet{chu2013}} & LEAs \\
Consecutive Matching Bullets Striae (CMS) &  &  0\% &  52\% \\ 
\end{tabular}}
\caption{\label{tab:bullets-ER} Error Rates in same-source land-to-land analysis reported in the literature.}
\end{table}

@@

In this paper we evaluate the adaptability of the Chumbley score as a
measure to quantify similarity in land engraved areas (LEAs) on bullets.
For that we briefly introduce the deterministic method suggested by
\citet{hadler} in section 2. In the process we provide methods to
identify parameters that minimize the error rates. We then investigate
persistent scenarios in which the method proposed by \citet{hadler}
fails to come to a result. We go on to provide a solution to the failed
tests problem, consequently increasing the power of the test and
reducing error rates in the process. We set up a testing framework to
compare the performance of the two algorithms in section 3 and finally
discuss results in section 4.
